---
title: "Testing hierarchical pathway kinetics with residue data on dimethenamid and dimethenamid-P"
author: Johannes Ranke
date: Last change on 20 April 2023, last compiled on `r format(Sys.time(), "%e %B %Y")`
geometry: margin=2cm
bibliography: references.bib
toc: true
output:
  pdf_document:
    extra_dependencies: ["float", "listing"]
---

```{r setup, echo = FALSE, cache = FALSE}
options(width = 80) # For summary listings
knitr::opts_chunk$set(
  comment = "", tidy = FALSE, cache = TRUE, fig.pos = "H", fig.align = "center"
)
```

\clearpage

# Introduction

The purpose of this document is to test demonstrate how nonlinear hierarchical
models (NLHM) based on the parent degradation models SFO, FOMC, DFOP and HS,
with parallel formation of two or more metabolites can be fitted with the mkin package.

It was assembled in the course of work package 1.2 of Project Number 173340
(Application of nonlinear hierarchical models to the kinetic evaluation of
chemical degradation data) of the German Environment Agency carried out in 2022
and 2023.

The mkin package is used in version `r packageVersion("mkin")`, which is currently
under development. It contains the test data, and the functions used in the
evaluations. The `saemix` package is used as a backend for fitting the NLHM,
but is also loaded to make the convergence plot function available.

This document is processed with the `knitr` package, which also provides the
`kable` function that is used to improve the display of tabular data in R
markdown documents. For parallel processing, the `parallel` package is used.

```{r, packages, cache = FALSE, message = FALSE}
library(mkin)
library(knitr)
library(saemix)
library(parallel)
n_cores <- detectCores()

# We need to start a new cluster after defining a compiled model that is
# saved as a DLL to the user directory, therefore we define a function
# This is used again after defining the pathway model
start_cluster <- function(n_cores) {
  if (Sys.info()["sysname"] == "Windows") {
    ret <- makePSOCKcluster(n_cores)
  } else {
    ret <- makeForkCluster(n_cores)
  }
  return(ret)
}
```

\clearpage

# Data

The test data are available in the mkin package as an object of class `mkindsg`
(mkin dataset group) under the identifier `dimethenamid_2018`. The following
preprocessing steps are done in this document.

- The data available for the enantiomer dimethenamid-P (DMTAP) are renamed
  to have the same substance name as the data for the racemic mixture
  dimethenamid (DMTA). The reason for this is that no difference between their
  degradation behaviour was identified in the EU risk assessment.
- Unnecessary columns are discarded
- The observation times of each dataset are multiplied with the
  corresponding normalisation factor also available in the dataset, in order to
  make it possible to describe all datasets with a single set of parameters
  that are independent of temperature
- Finally, datasets observed in the same soil (`Elliot 1` and `Elliot 2`) are
  combined, resulting in dimethenamid (DMTA) data from six soils.

The following commented R code performs this preprocessing.

```{r, data}
# Apply a function to each of the seven datasets in the mkindsg object to create a list
dmta_ds <- lapply(1:7, function(i) {
  ds_i <- dimethenamid_2018$ds[[i]]$data                     # Get a dataset
  ds_i[ds_i$name == "DMTAP", "name"] <-  "DMTA"              # Rename DMTAP to DMTA
  ds_i <- subset(ds_i, select = c("name", "time", "value")) # Select data
  ds_i$time <- ds_i$time * dimethenamid_2018$f_time_norm[i]  # Normalise time
  ds_i                                                       # Return the dataset
})

# Use dataset titles as names for the list elements
names(dmta_ds) <- sapply(dimethenamid_2018$ds, function(ds) ds$title)

# Combine data for Elliot soil to obtain a named list with six elements
dmta_ds[["Elliot"]] <- rbind(dmta_ds[["Elliot 1"]], dmta_ds[["Elliot 2"]]) #
dmta_ds[["Elliot 1"]] <- NULL
dmta_ds[["Elliot 2"]] <- NULL
```

\clearpage

The following tables show the `r length(dmta_ds)` datasets.

```{r show-data, dependson = "data", results = "asis"}
for (ds_name in names(dmta_ds)) {
  print(
    kable(mkin_long_to_wide(dmta_ds[[ds_name]]),
      caption = paste("Dataset", ds_name),
      booktabs = TRUE, row.names = FALSE))
    cat("\n\\clearpage\n")
}
```

# Separate evaluations

As a first step to obtain suitable starting parameters for the NLHM fits, we do
separate fits of several variants of the pathway model used previously
[@ranke2021], varying the kinetic model for the parent compound. Because the
SFORB model often provides faster convergence than the DFOP model, and can
sometimes be fitted where the DFOP model results in errors, it is included in
the set of parent models tested here.

```{r, sep-1-const, dependson = "data"}
if (!dir.exists("dmta_dlls")) dir.create("dmta_dlls")
m_sfo_path_1 <- mkinmod(
  DMTA = mkinsub("SFO", c("M23", "M27", "M31")),
  M23 = mkinsub("SFO"),
  M27 = mkinsub("SFO"),
  M31 = mkinsub("SFO", "M27", sink = FALSE),
  name = "m_sfo_path", dll_dir = "dmta_dlls",
  unload = TRUE, overwrite = TRUE,
  quiet = TRUE
)
m_fomc_path_1 <- mkinmod(
  DMTA = mkinsub("FOMC", c("M23", "M27", "M31")),
  M23 = mkinsub("SFO"),
  M27 = mkinsub("SFO"),
  M31 = mkinsub("SFO", "M27", sink = FALSE),
  name = "m_fomc_path", dll_dir = "dmta_dlls",
  unload = TRUE, overwrite = TRUE,
  quiet = TRUE
)
m_dfop_path_1 <- mkinmod(
  DMTA = mkinsub("DFOP", c("M23", "M27", "M31")),
  M23 = mkinsub("SFO"),
  M27 = mkinsub("SFO"),
  M31 = mkinsub("SFO", "M27", sink = FALSE),
  name = "m_dfop_path", dll_dir = "dmta_dlls",
  unload = TRUE, overwrite = TRUE,
  quiet = TRUE
)
m_sforb_path_1 <- mkinmod(
  DMTA = mkinsub("SFORB", c("M23", "M27", "M31")),
  M23 = mkinsub("SFO"),
  M27 = mkinsub("SFO"),
  M31 = mkinsub("SFO", "M27", sink = FALSE),
  name = "m_sforb_path", dll_dir = "dmta_dlls",
  unload = TRUE, overwrite = TRUE,
  quiet = TRUE
)
m_hs_path_1 <- mkinmod(
  DMTA = mkinsub("HS", c("M23", "M27", "M31")),
  M23 = mkinsub("SFO"),
  M27 = mkinsub("SFO"),
  M31 = mkinsub("SFO", "M27", sink = FALSE),
  name = "m_hs_path", dll_dir = "dmta_dlls",
  unload = TRUE, overwrite = TRUE,
  quiet = TRUE
)
cl <- start_cluster(n_cores)

deg_mods_1 <- list(
  sfo_path_1 = m_sfo_path_1,
  fomc_path_1 = m_fomc_path_1,
  dfop_path_1 = m_dfop_path_1,
  sforb_path_1 = m_sforb_path_1,
  hs_path_1 = m_hs_path_1)

sep_1_const <- mmkin(
  deg_mods_1,
  dmta_ds,
  error_model = "const",
  quiet = TRUE)

status(sep_1_const) |> kable()
```

All separate pathway fits with SFO or FOMC for the parent and constant variance
converged (status OK). Most fits with DFOP or SFORB for the parent converged
as well. The fits with HS for the parent did not converge with default settings.

```{r, sep-1-tc, dependson = "sep-1-const"}
sep_1_tc <- update(sep_1_const, error_model = "tc")
status(sep_1_tc) |> kable()
```

With the two-component error model, the set of fits with convergence problems
is slightly different, with convergence problems appearing for different data
sets when applying the DFOP and SFORB model and some additional convergence
problems when using the FOMC model for the parent.

\clearpage

# Hierarchichal model fits

The following code fits two sets of the corresponding hierarchical models to
the data, one assuming constant variance, and one assuming two-component error.

```{r saem-1, dependson = c("sep-1-const", "sep-1-tc")}
saem_1 <- mhmkin(list(sep_1_const, sep_1_tc))
```
The run time for these fits was around two hours on five year old hardware. After
a recent hardware upgrade these fits complete in less than twenty minutes.

```{r, saem-1-status, dependson = "saem-1"}
status(saem_1) |> kable()
```

According to the `status` function, all fits terminated successfully.

```{r saem-1-anova, dependson = "saem-1"}
anova(saem_1) |> kable(digits = 1)
```

When the goodness-of-fit of the models is compared, a warning is obtained,
indicating that the likelihood of the pathway fit with SFORB for the parent
compound and constant variance could not be calculated with importance sampling
(method 'is'). As this is the default method on which all AIC and BIC
comparisons are based, this variant is not included in the model comparison
table. Comparing the goodness-of-fit of the remaining models, HS model model
with two-component error provides the best fit. However, for batch experiments
performed with constant conditions such as the experiments evaluated here,
there is no reason to assume a discontinuity, so the SFORB model is
preferable from a mechanistic viewpoint. In addition, the information criteria
AIC and BIC are very similar for HS and SFORB. Therefore, the SFORB model is
selected here for further refinements.

\clearpage

## Parameter identifiability based on the Fisher Information Matrix

Using the `illparms` function, ill-defined statistical model parameters such as
standard deviations of the degradation parameters in the population and error
model parameters can be found.

```{r saem-1-illparms, dependson = "saem-1"}
illparms(saem_1) |> kable()
```

When using constant variance, no ill-defined variance parameters are identified
with the `illparms` function in any of the degradation models. When using
the two-component error model, there is one ill-defined variance parameter
in all variants except for the variant using DFOP for the parent compound.

For the selected combination of the SFORB pathway model with two-component
error, the random effect for the rate constant from reversibly bound DMTA to
the free DMTA (`k_DMTA_bound_free`) is not well-defined. Therefore, the fit is
updated without assuming a random effect for this parameter.

```{r saem-sforb-path-1-tc-reduced, dependson = "saem-1"}
saem_sforb_path_1_tc_reduced <- update(saem_1[["sforb_path_1", "tc"]],
  no_random_effect = "log_k_DMTA_bound_free")
illparms(saem_sforb_path_1_tc_reduced)
```

As expected, no ill-defined parameters remain. The model comparison below shows
that the reduced model is preferable.

```{r saem-sforb-path-1-tc-reduced-anova, dependson = "saem-sforb-path-1-tc-reduced"}
anova(saem_1[["sforb_path_1", "tc"]], saem_sforb_path_1_tc_reduced) |> kable(digits = 1)
```

The convergence plot of the refined fit is shown below.

```{r saem-sforb-path-1-tc-reduced-convergence, dependson = "saem-sforb-path-1-tc-reduced", fig.height = 12}
plot(saem_sforb_path_1_tc_reduced$so, plot.type = "convergence")
```

For some parameters, for example for `f_DMTA_ilr_1` and `f_DMTA_ilr_2`, i.e.
for two of the parameters determining the formation fractions of the parallel
formation of the three metabolites, some movement of the parameters is still
visible in the second phase of the algorithm. However, the amplitude of this
movement is in the range of the amplitude towards the end of the first phase.
Therefore, it is likely that an increase in iterations would not improve the
parameter estimates very much, and it is proposed that the fit is acceptable.
No numeric convergence criterion is implemented in saemix.

\clearpage

## Alternative check of parameter identifiability

As an alternative check of parameter identifiability [@duchesne_2021],
multistart runs were performed on the basis of the refined fit shown above.

```{r saem-sforb-multistart, dependson = "saem-sforb-path-1-tc-reduced"}
saem_sforb_path_1_tc_reduced_multi <- multistart(saem_sforb_path_1_tc_reduced,
  n = 32, cores = 10)
```

```{r dependson = "saem-sforb-multistart"}
print(saem_sforb_path_1_tc_reduced_multi)
```

Out of the 32 fits that were initiated, only 17 terminated without an error.
The reason for this is that the wide variation of starting parameters in combination
with the parameter variation that is used in the SAEM algorithm leads to
parameter combinations for the degradation model that the numerical integration
routine cannot cope with. Because of this variation of initial parameters,
some of the model fits take up to two times more time than the original fit.

```{r dependson = "saem-sforb-multistart", fig.cap = "Parameter boxplots for the multistart runs that succeeded", fig.height = 6, fig.width = 10}
par(mar = c(12.1, 4.1, 2.1, 2.1))
parplot(saem_sforb_path_1_tc_reduced_multi, ylim = c(0.5, 2), las = 2)
```

However, visual analysis of the boxplot of the parameters obtained in the
successful fits confirms that the results are sufficiently independent of the
starting parameters, and there are no remaining ill-defined parameters.

\clearpage


# Plots of selected fits

The SFORB pathway fits with full and reduced parameter distribution model are
shown below.

```{r fig.cap = "SFORB pathway fit with two-component error", dependson = "saem-1", fig.height = 8}
plot(saem_1[["sforb_path_1", "tc"]])
```

\clearpage

```{r fig.cap = "SFORB pathway fit with two-component error, reduced parameter model", dependson = "saem-sforb-path-1-tc-reduced", fig.height = 8}
plot(saem_sforb_path_1_tc_reduced)
```

Plots of the remaining fits and listings for all successful fits are shown in
the Appendix.

```{r}
stopCluster(cl)
```


# Conclusions

Pathway fits with SFO, FOMC, DFOP, SFORB and HS models for the parent compound
could be successfully performed.

\clearpage

# Acknowledgements

The helpful comments by Janina Wöltjen of the German Environment Agency
on earlier versions of this document are gratefully acknowledged.

# References

\vspace{1em}

::: {#refs}
:::

\clearpage

# Appendix

## Plots of hierarchical fits not selected for refinement

```{r fig.cap = "SFO pathway fit with two-component error", dependson = "saem-1", fig.height = 8}
plot(saem_1[["sfo_path_1", "tc"]])
```

\clearpage

```{r fig.cap = "FOMC pathway fit with two-component error", dependson = "saem-1", fig.height = 8}
plot(saem_1[["fomc_path_1", "tc"]])
```

\clearpage


```{r fig.cap = "HS pathway fit with two-component error", dependson = "saem-1", fig.height = 8}
plot(saem_1[["sforb_path_1", "tc"]])
```

\clearpage

## Hierarchical model fit listings

### Fits with random effects for all degradation parameters

```{r listings-1, results = "asis", echo = FALSE}
errmods <- c(const = "constant variance", tc = "two-component error")
degmods <- c(
  sfo_path_1 = "SFO path 1",
  fomc_path_1 = "FOMC path 1",
  dfop_path_1 = "DFOP path 1",
  sforb_path_1 = "SFORB path 1",
  hs_path_1 = "HS path 1")
for (deg_mod in rownames(saem_1)) {
  for (err_mod in c("const", "tc")) {
    fit <- saem_1[[deg_mod, err_mod]]
    if (!inherits(fit$so, "try-error")) {
      caption <- paste("Hierarchical", degmods[deg_mod], "fit with", errmods[err_mod])
      tex_listing(fit, caption)
    }
  }
}
```

### Improved fit of the SFORB pathway model with two-component error

```{r listings-2, results = "asis", echo = FALSE, dependson = "listings-1"}
caption <- paste("Hierarchical SFORB pathway fit with two-component error")
tex_listing(saem_sforb_path_1_tc_reduced, caption)
```

## Session info

```{r, echo = FALSE}
sessionInfo()
```

## Hardware info

```{r, echo = FALSE}
if(!inherits(try(cpuinfo <- readLines("/proc/cpuinfo")), "try-error")) {
  cat(gsub("model name\t: ", "CPU model: ", cpuinfo[grep("model name", cpuinfo)[1]]))
}
if(!inherits(try(meminfo <- readLines("/proc/meminfo")), "try-error")) {
  cat(gsub("model name\t: ", "System memory: ", meminfo[grep("MemTotal", meminfo)[1]]))
}
```
