---
title: "Benchmark timings for mkin"
author: "Johannes Ranke"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    fig_retina: null
date: "`r Sys.Date()`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(knitr)
opts_chunk$set(tidy = FALSE, cache = FALSE)
library("mkin")
```

Each system is characterized by its CPU type, the operating system type and the
mkin version. Currently only values for one system are available.

```{r}
cpu_model <- benchmarkme::get_cpu()$model_name
operating_system <- Sys.info()[["sysname"]]
mkin_version <- as.character(packageVersion("mkin"))
system_string <- paste0(operating_system, ", ", cpu_model, ", mkin version ", mkin_version)
load("~/git/mkin/vignettes/web_only/mkin_benchmarks.rda")
mkin_benchmarks[system_string, c("CPU", "OS", "mkin")] <-
  c(cpu_model, operating_system, mkin_version)

if (mkin_version > "0.9.48.1") {
  mmkin_bench <- function(models, datasets, error_model = "const") {
    mmkin(models, datasets, error_model = error_model, cores = 1, quiet = TRUE)
  }
} else {
  mmkin_bench <- function(models, datasets, error_model = NULL) {
    mmkin(models, datasets, reweight.method = error_model, cores = 1, quiet = TRUE)
  }
}
```

## Test cases

Parent only:

```{r parent_only, warning = FALSE}
FOCUS_C <- FOCUS_2006_C
FOCUS_D <- subset(FOCUS_2006_D, value != 0)
parent_datasets <- list(FOCUS_C, FOCUS_D)

t1 <- system.time(mmkin_bench(c("SFO", "FOMC", "DFOP", "HS"), parent_datasets))[["elapsed"]]
t2 <- system.time(mmkin_bench(c("SFO", "FOMC", "DFOP", "HS"), parent_datasets,
    error_model = "tc"))[["elapsed"]]
```

One metabolite:

```{r one_metabolite}
SFO_SFO <- mkinmod(
  parent = mkinsub("SFO", "m1"),
  m1 = mkinsub("SFO"))
FOMC_SFO <- mkinmod(
  parent = mkinsub("FOMC", "m1"),
  m1 = mkinsub("SFO"))
DFOP_SFO <- mkinmod(
  parent = mkinsub("FOMC", "m1"),
  m1 = mkinsub("SFO"))
t3 <- system.time(mmkin_bench(list(SFO_SFO, FOMC_SFO, DFOP_SFO), list(FOCUS_D)))[["elapsed"]]
t4 <- system.time(mmkin_bench(list(SFO_SFO, FOMC_SFO, DFOP_SFO), list(FOCUS_D),
    error_model = "tc"))[["elapsed"]]
t5 <- system.time(mmkin_bench(list(SFO_SFO, FOMC_SFO, DFOP_SFO), list(FOCUS_D),
    error_model = "obs"))[["elapsed"]]
```

Two metabolites, synthetic data:

```{r two_metabolites}
m_synth_SFO_lin <- mkinmod(parent = mkinsub("SFO", "M1"),
                           M1 = mkinsub("SFO", "M2"),
                           M2 = mkinsub("SFO"),
                           use_of_ff = "max", quiet = TRUE)

m_synth_DFOP_par <- mkinmod(parent = mkinsub("DFOP", c("M1", "M2")),
                           M1 = mkinsub("SFO"),
                           M2 = mkinsub("SFO"),
                           use_of_ff = "max", quiet = TRUE)

SFO_lin_a <- synthetic_data_for_UBA_2014[[1]]$data

DFOP_par_c <- synthetic_data_for_UBA_2014[[12]]$data

t6 <- system.time(mmkin_bench(list(m_synth_SFO_lin), list(SFO_lin_a)))["elapsed"]
t7 <- system.time(mmkin_bench(list(m_synth_DFOP_par), list(DFOP_par_c)))["elapsed"]

t8 <- system.time(mmkin_bench(list(m_synth_SFO_lin), list(SFO_lin_a),
    error_model = "tc"))["elapsed"]
t9 <- system.time(mmkin_bench(list(m_synth_DFOP_par), list(DFOP_par_c),
    error_model = "tc"))["elapsed"]

t10 <- system.time(mmkin_bench(list(m_synth_SFO_lin), list(SFO_lin_a),
    error_model = "obs"))["elapsed"]
t11 <- system.time(mmkin_bench(list(m_synth_DFOP_par), list(DFOP_par_c),
    error_model = "obs"))["elapsed"]
```

```{r results}
mkin_benchmarks[system_string, paste0("t", 1:11)] <-
  c(t1, t2, t3, t4, t5, t6, t7, t8, t9, t10, t11)
save(mkin_benchmarks, file = "~/git/mkin/vignettes/web_only/mkin_benchmarks.rda")
```

## Results

Currently, we only have benchmark information on one system, therefore only the mkin
version is shown with the results below. Timings are in seconds, shorter is better.

Benchmarks for all available error models are shown.

### Parent only

Constant variance and two-component error model:

```{r}
print(mkin_benchmarks[, c("mkin", "t1", "t2")], row.names = FALSE)
```

### One metabolite

Constant variance, variance by variable and two-component error model:

```{r}
print(mkin_benchmarks[, c("mkin", "t3", "t4", "t5")], row.names = FALSE)
```

### Two metabolites

Two different datasets, for each constant variance, variance by variable and
two-component error model are shown:

```{r}
print(mkin_benchmarks[, c("mkin", paste0("t", 6:11))], row.names = FALSE)
```
